Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-05 paper/main.tex
some tweaks, initial add of first appendix

We next place a zero-mean Gaussian process prior on $S$, i.e.,
$S\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_S(\cdot)\right)$, again with a 
spatial kernel.  
\textcolor{red}{(@Steve - is a zero mean the right thing?  If not, what is the mean?)}
Standard practice would be to use locations $X$ as the inputs to kernel $\Sigma_S(\cdot)$.
Yet we have found that prior to lack sufficient flexibility for our motivating
application.  In the Mira-Titan setting, some nonstationarity is expected due to 
baryonic acoustic oscillations when $-2 \leq \log_{10}(k) \leq -1$. To account 
for this, we incorporate an additional latent Gaussian process, $W$, which accommodates
nonstationarity by warping $X$.  Ultimately, this yields the following hierarchical model:
\begin{equation}\label{eq:dgphm}
\begin{aligned}
\bar{Y}\mid S,\Sigma_\varepsilon &\sim \mathrm{GP}(S,\Sigma_\varepsilon) \\
S\mid W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_S(W)\right) \\
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{aligned}
\end{equation}
$W$ is also an $n$-vector whose entries $w_i$ correspond to warped versions of each $x_i$.
The functional composition of GPs in the latter two lines forms a ``deep Gaussian process'' 
\citep{damianou2013deep,dunlop2018deep}. Additional latent layers could be considered, 
although we find one is sufficient.

@@

While our inferential goal is to obtain the posterior distribution of 
$S\mid\bar{Y},\Sigma_\varepsilon$, this requires inferring latent quantity $W$.  
We first integrate over $S$ to condense our hierarchical model of Eq.~\ref{eq:dgphm} into
\begin{align}
\label{eq:likelihood}
\bar{Y} \mid W, \Sigma_\varepsilon &\sim \textrm{GP}(\mathbf{0}, \Sigma_S(W) + \Sigma_\varepsilon) \\
\label{eq:prior}
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{align}
Details are provided in Appendix \textcolor{red}{???}. We then use elliptical slice
sampling \citep[ESS;][]{murray2010elliptical} to generate posterior samples of $W$.  
Proposals are generated from prior samples from Eq.~(\ref{eq:prior}), and acceptances 
involve the ratio of likelihoods from Eq.~\ref{eq:likelihood}.

@@

Given burned-in posterior samples $\{W^{(t)}, \tau^{2(t)}, \theta_S^{(t)}, \theta_W^{(t)}\}$
for $t\in\mathcal{T}$, we may obtain the
posterior distribution of the true power matter spectra for a particular cosmology as
\[
S^{(t)}\mid\bar{Y}, \Sigma_\varepsilon \sim \mathcal{N}_n(m^{(t)}, C^{(t)})
\quad\textrm{where}\quad
\begin{array}{rl}
C^{(t)}&=\left(\Sigma_S(W^{(t)})^{-1}+\Sigma_\varepsilon^{-1}\right)^{-1} \\
m^{(t)}&=C^{(t)}\left(\Sigma_\varepsilon^{-1}\bar{Y}+\Sigma_S(W^{(t)})^{-1}\mu\right).
\end{array}
\] 
\textcolor{red}{(@Steve - provide details of this derivation.  Also what is $\mu$?  Are we 
going with zero here?  Add details of how we combine these things across all the $t$ samples.)}
We denote our inference procedure for the hierarchical model of Eq.~\ref{eq:dgphm} as
``DGP.HM'' given its deep Gaussian process foundation with additional hierarchical modeling (HM)
structure to incorporate multiple functional outputs.
\textcolor{red}{(Is it too late to change this?  Maybe DGP-FCO for functional correlated output?)}

@@

\section{Appendix}
\label{sec:apdx}

>>>>

We next place a Gaussian process prior on $S$, i.e.,
$S\sim \mathrm{GP}\left(\boldsymbol{\mu}, \Sigma_S(\cdot)\right)$, again with a 
spatial kernel. Here, we set $\boldsymbol{\mu}=\mathbf{0}$, although the modeling can 
be carried out with any selection of $\boldsymbol{\mu}$. Standard practice would 
be to use locations $X$ as the inputs to kernel $\Sigma_S(\cdot)$.
Yet we have found that prior to lack sufficient flexibility for our motivating
application.  In the Mira-Titan setting, some nonstationarity is expected due to 
baryonic acoustic oscillations when $-2 \leq \log_{10}(k) \leq -1$. To account 
for this, we incorporate an additional latent Gaussian process, $W$, which accommodates
nonstationarity by warping $X$.  Ultimately, this yields the following hierarchical model:
\begin{equation}\label{eq:dgphm}
\begin{aligned}
\bar{Y}\mid S,\Sigma_\varepsilon &\sim \mathrm{GP}(S,\Sigma_\varepsilon) \\
S\mid W &\sim \mathrm{GP}\left(\boldsymbol{\mu}, \Sigma_S(W)\right) \\
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{aligned}
\end{equation}
$W$ is also an $n$-vector whose entries $w_i$ correspond to warped versions of each $x_i$.
The functional composition of GPs in the latter two lines forms a ``deep Gaussian process'' 
\citep{damianou2013deep,dunlop2018deep}. Additional latent layers could be considered, 
although we find one is sufficient.

@@

While our inferential goal is to obtain the posterior distribution of 
$S\mid\bar{Y},\Sigma_\varepsilon$, this requires inferring latent quantity $W$.  
We first integrate over $S$ to condense our hierarchical model of Eq.~\ref{eq:dgphm} into
\begin{align}
\label{eq:likelihood}
\bar{Y} \mid W, \Sigma_\varepsilon &\sim \textrm{GP}(\boldsymbol{\mu}, \Sigma_S(W) + \Sigma_\varepsilon) \\
\label{eq:prior}
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{align}
Details are provided in Appendix \textcolor{red}{???}. We then use elliptical slice
sampling \citep[ESS;][]{murray2010elliptical} to generate posterior samples of $W$.  
Proposals are generated from prior samples from Eq.~(\ref{eq:prior}), and acceptances 
involve the ratio of likelihoods from Eq.~\ref{eq:likelihood}.

@@

Given burned-in posterior samples $\{W^{(t)}, \tau^{2(t)}, \theta_S^{(t)}, \theta_W^{(t)}\}$
for $t\in\mathcal{T}$, we may obtain the
posterior distribution of the true power matter spectra for a particular cosmology as
\[
S^{(t)}\mid\bar{Y}, \Sigma_\varepsilon \sim \mathcal{N}_n(m^{(t)}, C^{(t)})
\quad\textrm{where}\quad
\begin{array}{rl}
C^{(t)}&=\left(\Sigma_S(W^{(t)})^{-1}+\Sigma_\varepsilon^{-1}\right)^{-1} \\
m^{(t)}&=C^{(t)}\left(\Sigma_\varepsilon^{-1}\bar{Y}+\Sigma_S(W^{(t)})^{-1}\boldsymbol{\mu}\right).
\end{array}
\] 
\textcolor{red}{(@Steve - provide details of this derivation.  Also what is $\mu$?  Are we 
going with zero here?  Add details of how we combine these things across all the $t$ samples.)}
We denote our inference procedure for the hierarchical model of Eq.~\ref{eq:dgphm} as
``DGP.HM'' given its deep Gaussian process foundation with additional hierarchical modeling (HM)
structure to incorporate multiple functional outputs.
\textcolor{red}{(Is it too late to change this?  Maybe DGP-FCO for functional correlated output?)}

@@

\appendix
\renewcommand{\thesection}{Appendix \Alph{section}}

\section{Integrated Likelihood}
\label{sec:apdx_int_lik}
Here, we establish that under the model:

$$
\begin{aligned}
Y_i|S,W &\sim N\big(S, \Sigma_\varepsilon(w,w')\big), \quad i \in \{1,\dots, r\}\\
S|W &\sim N\big(m, \Sigma_S(w,w')\big)\\
W &\sim N\big(0, \Sigma_W(x,x')\big) \\
\end{aligned}
$$

We can obtain the integrated likelihood $p(Y_i|W)=\int p(Y_i|S,W)p(S|W) dS$, with the following calculations:
$$
\begin{aligned}
p(Y_i|W) &= \int p(Y_i|S,W)p(S|W)dS \\
&\propto \int\exp\left(-\frac{1}{2} \left[Y_i^T\Sigma_\varepsilon^{-1} Y_i - 2Y_i^T\Sigma_\varepsilon^{-1} S + S^T \Sigma_\varepsilon^{-1} S + S^T \Sigma_S^{-1} S -2S^T \Sigma_S^{-1} m + m^T \Sigma_S^{-1} m \right]\right)dS \\
\end{aligned}
$$

If we focus only on terms containing $S$, we have 
$$
\begin{aligned}
    \exp\Big(-\frac{1}{2} \big[S^T &(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1}) S -2S^T (\Sigma_S^{-1} m + \Sigma_\varepsilon^{-1} Y_i) \big]\Big) \\
&= \exp\left(-\frac{1}{2} \left[S^T (\Sigma_\varepsilon^{-1}+\Sigma_S^{-1}) S -2S^T (\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}(\Sigma_S^{-1} m + \Sigma_\varepsilon^{-1} Y_i) \right]\right)
\end{aligned}
$$

To complete the square, we add and subtract $(\Sigma_S^{-1} m + \Sigma_\varepsilon^{-1} Y_i)^T(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}(\Sigma_S^{-1} m + \Sigma_\varepsilon^{-1} Y_i)$, which then gives the kernel for $S \sim N\left((\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}(\Sigma_S^{-1} m + \Sigma_\varepsilon^{-1} Y_i), (\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1} \right)$. This removes the integral and leaves us with the following terms for $Y_i|W$:
$$
\begin{aligned}
&\exp\Big(-\frac{1}{2}\big[Y_i^T\Sigma_\varepsilon^{-1} Y + m^T\Sigma_S^{-1} m - (\Sigma_S^{-1} m + \Sigma_\varepsilon^{-1} Y_i)^T(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1} (\Sigma_S^{-1} m + \Sigma_\varepsilon^{-1} Y_i)\big]\Big) \\
&= \exp\Big(-\frac{1}{2}\big[Y_i^T\big(\Sigma_\varepsilon^{-1} - \Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}\Sigma_\varepsilon^{-1}\big)Y_i -2Y_i^T \Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}\Sigma_S^{-1} m \\
&\qquad \qquad \qquad + m^T (\Sigma_S^{-1} - \Sigma_S^{-1}(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}\Sigma_S^{-1}) m\big]\Big) \\
&= \exp\Big(-\frac{1}{2}\big((Y_i-m)^T(\Sigma_\varepsilon+\Sigma_S)^{-1}(Y_i-m) \big)\Big)
\end{aligned}
$$

With the last equality resulting from the use of the Sherman–Morrison–Woodbury formula, yielding $\big(\Sigma_\varepsilon^{-1} - \Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}\Sigma_\varepsilon^{-1} = (\Sigma_S^{-1} - \Sigma_S^{-1}(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}\Sigma_S^{-1}) = (\Sigma_\varepsilon+\Sigma_S)^{-1}$. We also use the fact that $\Sigma_\varepsilon^{-1}(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})^{-1}\Sigma_S^{-1} = (\Sigma_\varepsilon + \Sigma_S)^{-1}$, which holds since it can be shown that $[\Sigma_S(\Sigma_\varepsilon^{-1}+\Sigma_S^{-1})\Sigma_\varepsilon](\Sigma_\varepsilon+\Sigma_S)=(\Sigma_\varepsilon+\Sigma_S)(\Sigma_\varepsilon+\Sigma_S)$.

Therefore, we have that $Y_i|W \sim N\big(m, \Sigma_S(w,w')+\Sigma_\varepsilon(w,w')\big)$.

\section{Mira-Titan Predictive Plots}
\label{sec:apdx_mira_plots}
